---
title: "Reproducible Research: Peer Assessment 1"
author: "Chan Han Chin"
date: "March 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing data

Download the zip file to working directory and unzip it.

```{r}
if (!file.exists("activity.csv")){
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, "getdata.zip")
  unzip("getdata.zip")
}
```
Read the `activity.csv` file as `activity`. Check the structure of `activity`

```{r}
activity <- read.csv("activity.csv")
str(activity)
```

## What is the mean total number of steps taken per day?

Load the dplyr library for data manipulation and ggplot2 for plotting.
```{r, message = FALSE}
library(dplyr)
library(ggplot2)
```

1. Calculate the total number of steps taken per day.

```{r}
steps_df <- activity %>% group_by(date) %>% summarise(sum_steps = sum(steps, na.rm=TRUE))
```
2. Plot a histogram of total number of steps taken per day.

```{r, message = FALSE}
main <- ggplot(steps_df, aes(x=sum_steps)) + stat_bin(fill="blue") + theme_bw()
lab <- labs(title = " Total Steps Taken per Day", x = "Total Steps", y = "Days") 
hist_plot <- main + lab + theme(plot.title = element_text(hjust = 0.5)) + scale_y_continuous(breaks = c(0,2,4,6,8,10))
hist_plot
```

3. Calculate the mean and median of total number of steps taken per day.
```{r}
summary(steps_df)
```

* The mean and median of total number of steps taken per day are 9354 and 10395.

## What is the average daily activity pattern?

1. Plot the time series of average number of steps taken across all days (y-axis) for each 5 minute interval (x-axis)

* First, create a new dataframe `avg_steps_df` which gives the average steps across all days for each interval.

```{r}
avg_steps_df <- activity %>% group_by(interval) %>% summarise(avg_steps = mean(steps, na.rm=TRUE))
```

* Then, use gpplot to draw the timeseries.

```{r}
main1 <- ggplot(avg_steps_df, aes(x=interval, y=avg_steps)) + geom_line() + theme_bw()
lab1 <- labs(title = "Average Steps For Each 5-Minute Interval", x = "Interval", y="Average Steps")
time_series_plot <- main1 + lab1 + theme(plot.title = element_text(hjust = 0.5))
time_series_plot
```

2. Which 5-minute interval averaged across all days contains the maximum number of steps?

```{r}
avg_steps_df[which.max(avg_steps_df$avg_steps),]
```

* Interval 835 contains the maximum number of steps on average, which is 206.17 steps.

## Imputing missing values

1. Calculate and report the total number of missing values (i.e NA) in the dataset.

```{r}
sum(is.na(activity))
```

* The total number of missing values is 2304.

2. Devise a strategy to fill in all missing values in the dataset.  

* The previous time series plot suggests that for certain intervals, the average number of steps is consistently low - not more than 10 steps for the intervals from 2200 to 530 (i.e 10pm to 530am).The average number of steps starts to increase around 600 (6am). Apart from the big spike from less than 100 steps to around 200 steps for the intervals of 800 (8am) to 900 (9am), the average number of steps  oscilattes around 25 steps to 100 steps for each interval througout the day till 2000 (10pm). 

* The above suggests that, the individual has a consistent sleeping pattern after 10pm and wakes up after 530am. Hence, for this assignment, I decided to use mean number of steps across all days for each time interval to impute the missing values in the dataset.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

* Merge the `activity` dataset with the `avg_steps_df` dataset by interval.An `activity2` dataframe is created.

```{r}
activity2 <- merge(activity, avg_steps_df, by="interval")
```

* Replace the missing steps with the average steps across all days for each time interval.

```{r}
activity2$steps[which(is.na(activity2$steps))] <- activity2$avg_steps[which(is.na(activity2$steps))]
```

* Do a simple check if there are any more missing values.
```{r}
sum(is.na(activity2))
```

* No more missing values.

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

* Create dataframe `steps_df2` which stores the new total number of steps per day information after imputing for missing values.

```{r}
steps_df2 <- activity2 %>% group_by(date) %>% summarise(sum_steps = sum(steps))
```

* Use ggplot to make the histogram of total number of steps taken each day.

```{r, message=FALSE}
main2 <- ggplot(steps_df2, aes(x=sum_steps)) + stat_bin(fill="red") + theme_bw()
lab2 <- labs(title = "Total Steps per Day after Imputation", x = "Total Steps", y = "Days")
hist_plot2 <- main2 + lab2 + theme(plot.title = element_text(hjust=0.5)) + scale_y_continuous(breaks=c(0,2,4,6,8,10,12,14))
hist_plot2
```

* Calculate mean and median for total number of steps taken per day.

```{r}
summary(steps_df2)
```

* The mean and median for total number of steps taken per day are now 10766 and 10766. These values have increased compared to the previous estimates. It seems that the imputation of missing values have the greatest impact on the first quartile, where the first quartile increased from 6778 to 9819. The distribution of the total number of steps taken per day has become more centered.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend".

* In order to do this, need to first convert the `date` variable (currently as factors) to date type. I use the lubridate package for easy conversion. Load the lubridate package then convert the `date` variable to date type. Then, create a `day` variable using `weekdays()`. Finally convert this `day` variable into a factor variable using `as.factor()`.

```{r, message = FALSE}
library(lubridate)
activity2$date <- ymd(activity2$date)
activity2$day <- ifelse(weekdays(activity2$date)=="Saturday" | weekdays(activity2$date)=="Sunday","weekend","weekday")
activity2$day <- as.factor(activity2$day)                       
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

* Create a new dataframe `activity_all` which summarizes the average steps by interval and day.

```{r}
activity_all <- activity2 %>% group_by(interval, day) %>% summarise(avg_steps = mean(steps))
```

* To draw 2 plots, I separate them into two separate dataframes again.

```{r}
activity_weekday <- activity_all %>% filter(day == "weekday")
activity_weekend <- activity_all %>% filter(day == "weekend")
```

* Load the package `gridExtra` to draw panel plots using ggplot. Use ggplot to plot the timeseries.

```{r, message = FALSE}
library(gridExtra)
first_main <- ggplot(activity_weekday, aes(x=interval, y = avg_steps)) + geom_line()
first_plot <- first_main + labs(title = "Weekday Acitivity", y = "Average Steps") + theme_bw()
second_main <- ggplot(activity_weekend, aes(x=interval, y = avg_steps)) + geom_line()
second_plot <- second_main + labs(title = "Weekend Activity", y = "Average Steps") + scale_y_continuous(limit = c(0, 200)) + theme_bw()
grid.arrange(first_plot, second_plot, nrow = 2)
```

* It can be seen that, the individual is more active early morning during weekdays (i.e activity starts about 530am. During weekends, the individual has a low average steps until about 8am, suggesting that the individual may spend more time in the bed during weekends. During the afternoon, the individual is more active on weekends compared to weekdays.